Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-09 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Fix overlapping affiliations

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands 
  - name: Gerko Vink
    affiliation: Utrecht University 
  - name: Valentijn M. T. de Jong
    affiliation: Julius Center for Health Sciences and Primary Care, University Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands \AND Data Analytics and Methods Task Force, European Medicines Agency, Amsterdam, The Netherlands

@@

```{r, setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(prompt = 'R> ', continue = '+ ')

@@

When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. [TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing. [TODO: Refer to Figure 1 and put interpretation in the figure caption.]

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  coord_cartesian(expand = c(0,0,0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on unfortunate outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
```

@@

As we can see... [TODO: fill in.]

@@

R=region+language+(1|InterviewID) model with with interviewer as random effects, because the observations are not independent. Interviews are not allocated randomly. In theory we expect the inclusion-restriction variable to be randomly assigned, that's why we're adding region and language to compensate for non-random allocation.

@@

# Categorical variables have to be recoded as factor
# to use the 2l.binary imputation method, it is required that the level names should not have any space between words

@@

Impute the data:
```{r}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

@@

>>>>

---
documentclass: jss
author:
  - name: Hanne I. Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Munoz Avila"
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands \AND
  - name: Thomas P. A. Debray
    affiliation: Julius Center for Health Sciences and Primary Care
    address: |
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands   
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Valentijn M. T. de Jong
    affiliation: |
      | Julius Center for Health Sciences and Primary Care, 
      | University Medical Center Utrecht, Utrecht University, 
      | Utrecht, The Netherlands 
    affiliation2: |
      | Data Analytics and Methods Task Force, 
      | European Medicines Agency, Amsterdam, The Netherlands

@@

```{r, setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(prompt = 'R> ', continue = '+ ')

@@

When clustered datasets are affected by missing values, we can distinguish between two types of missing data: sporadic missingness and systematic missingness [@resc13]. Sporadic missingness arises when variables are missing for some but not all of the units in a cluster [@buur18; @jola18]. For example, it is possible that test results are missing for several students in one or more classes. [TODO: Provide an example for one of the case studies below.] When all observations are missing within one or more clusters, data are systematically missing. [TODO: Refer to Figure 1 and put interpretation in the figure caption. VMTdJ: "Mention in text that this is sporadic missingness?"]

@@

```{r patterns, fig.height=1.5, fig.width=4.1, fig.cap = "Missingness in multilevel data", echo=FALSE}
dat <- expand.grid(rows = 1:7, cols = 1:6) %>% 
  cbind(text = c("1", "1", "2", "2", "3", "", "N", rep("", 35)),
        miss = c(rep("", 16), "NA", "NA", "", "", "", "NA", "", "", "NA", rep("", 17)))
ggplot(dat, aes(x = cols, y = rows)) +
  geom_tile(fill = "white", color = "black", size = 0.5) +
  geom_text(aes(label = text), color = "black", size = 3) +
  geom_text(aes(label = miss), color = mice:::mdc(2), family = "mono", fontface = "bold") + 
  scale_x_continuous(
    breaks = 1:6,
    labels = c("cluster", expression(X[1]), expression(X[2]), expression(X[3]), "...", expression(X[p])),
    name = NULL,
    position = "top"
  ) +
  scale_y_continuous(breaks = 1:7, labels = c(1:5, "...", "n"), name = NULL, trans = "reverse") +
  # coord_cartesian(expand = c(0,0)) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(0,0,0,0), "pt"))
# TODO: use math font for n and N
```
<!-- Column $X_1$ in Figure 1 is completely observed, column $X_2$ is systematically missing in cluster 2, and column $X_3$ is sporadically missing. To analyze these incomplete data, we have to take the nature of the missingness and the cluster structure into account. For example, the sporadic missingness in $X_3$ could be easily amended if this would be a cluster-level variable (and thus constant within clusters). We could then just extrapolate the true (but missing) value of $X_3$ for unit 1 from unit 2, and the value for unit 4 from unit 3. If $X_3$ would instead be a unit-level variable (which may vary within clusters), we could not just recover the unobserved 'truth', but would need to use some kind of missing data method, or discard the incomplete units altogether (i.e., complete case analysis). Complete case analysis can however introduce bias in statistical inferences and lowers statistical power. Further, with the systematic missingness in $X_2$, it would be impossible to fit a multilevel model without accommodating the missingness in some way. Complete case analysis in that case would mean excluding the entire cluster from the analyses. The wrong choice of missing data handling method can thus be extremely harmful to the inferences.  -->

@@

The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil` into the analysis model: 
```{r mod}
mod <- mort ~ 1 + type + age + motor_score + pupil + ct + (1 | name) 
```

@@

As we can see... [TODO: fill in: This means that a higher ct (Marshall Computerized Tomography classification) is associated with a lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI ... to ..., when controlling for ...]

@@

R=region+language+(1|InterviewID) model with interviewer as random effects, because the observations are not independent. Interviews are not allocated randomly. In theory we expect the inclusion-restriction variable to be randomly assigned, that's why we're adding region and language to compensate for non-random allocation. [TODO: Explain exclusion-restriction, refer to paper Johanna].

@@

# Categorical variables have to be recoded as factor
# to use the 2l.binary imputation method, it is required 
# that the level names should not have any space between words

@@

Impute the data:
```{r eval=FALSE}
imp <- mice( hivdata, # dataset with missing values
                        m = 10,   # number of imputations
                        maxit = 1,
                        seed = 1234, #seed attached to the dataID
                        meth = meth, #imputation method vector
                        pred = pred, #imputation predictors matrix
                        print = T,
                        meta_method="reml",
                        pmm=FALSE)
```

@@

# Funding

This project has received funding from the European Union's Horizon 2020 research and innovation programme under ReCoDID grant agreement No 825746.

The views expressed in this paper are the personal views of the authors and may not be understood or quoted as being made on behalf of or reflecting the position of the regulatory agency/agencies or organizations with which the authors are employed/affiliated.
